---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### NOTE: Figures are in "PA1_template_files/figure-html/"

## Loading and preprocessing the data
```{r activity}
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

```{r steps_per_day}
steps_per_day <- tapply(activity$steps[!is.na(activity$steps)],activity$date[!is.na(activity$steps)],FUN=sum)
hist(steps_per_day,col="red",
     main="Total number steps taken per day",
     xlab="steps per day")
mean_steps <- mean(steps_per_day[!is.na(steps_per_day)])
median_steps <- median(steps_per_day[!is.na(steps_per_day)])
```

The mean and median of total number of steps taken per day are 
`r as.integer(mean_steps)` and `r median_steps` respectively.

## What is the average daily activity pattern?

```{r steps_per_interval}
intervals<- as.factor(activity$interval[!is.na(activity$steps)])
steps_per_interval <- tapply(activity$steps[!is.na(activity$steps)],
                             intervals,mean)
plot(levels(intervals),steps_per_interval,type="l",col="blue",
     xlab="interval", ylab="average steps",main="Average Daily Activity")

max_activiy_interval <- names(which.max(steps_per_interval))
```

The interval `r max_activiy_interval` on average contains the maximum number of steps.

## Imputing missing values
```{r num_missing}
num_missing <- sum(is.na(activity$steps))
```

There are `r num_missing` missing values for steps in the activity data.

We impute the missing values with the average number of steps for the corresponding interval and create a new data set names `complete_activity`.

```{r complete_activity}
# New data set
complete_activity <- activity
complete_activity$steps[is.na(activity$steps)]  <- sapply(activity$interval[is.na(activity$steps)], 
                                                        function(x,steps_per_interval)
                                                        steps_per_interval[toString(x)],
                                                        steps_per_interval)

```

```{r new_steps_per_day}
# Average number of steps per day
new_steps_per_day <- tapply(complete_activity$steps,complete_activity$date,FUN=sum)
hist(new_steps_per_day,col="red",
     main="Total number steps taken per day for imputed data",
     xlab="steps per day")
new_mean_steps <- mean(new_steps_per_day)
new_median_steps <- median(new_steps_per_day)
```

The mean and median of total number of steps taken per day are 
`r as.integer(new_mean_steps)` and `r as.integer(new_median_steps)` respectively. These values differ from 
previously computed values from original data. Imputing missing values has improved the estimates for 
total number of steps per day, as now mean and median coincide. 



## Are there differences in activity patterns between weekdays and weekends?
```{r complete_activity$day}
complete_activity$day <- weekdays(as.POSIXlt(complete_activity$date))
complete_activity$day <- sapply(complete_activity$day,function(x) 
                                                        if (x=="Saturday" | x== "Sunday"){
                                                                return("weekend")
                                                        }
                                                        else {
                                                                return("weekday")
                                                        }
        
)
```
```{r weekday_weekend}
intervals<- complete_activity$interval
num_int <- length(unique(intervals))

pdat <- data.frame(rep(unique(intervals),2))
names(pdat) <- "interval"
 
pdat$day <- rep("weekend",2*num_int)

pdat$steps <- rep(0,2*num_int)

pdat$steps[1:num_int] <-tapply(complete_activity$steps[complete_activity$day=="weekend"],
                               intervals[complete_activity$day=="weekend"],
                               mean)

pdat$day[(num_int+1):(2*num_int)] <- "weekday"
#pdat$day <- as.factor(pdat$day)
pdat$steps[(num_int+1):(2*num_int)] <-
        tapply(complete_activity$steps[complete_activity$day=="weekday"],
                        intervals[complete_activity$day=="weekday"],
                        mean)

library(lattice)
xyplot(steps~interval | day, data = pdat,
     type = 'l',
      xlab = 'Interval',
      ylab = 'Number of Steps',
      layout = c(1,2))

```

On Weekdays the person is equally active through the daytime, whereas on weekends there is high level of activity(maybe he/she excercises) in late morning and then it decreases sharply (maybe watching movies).
